Haul-out behavior and aerial survey detectability of seals in the Bering and Chukchi seas

### 2021-03-29 analysis/paper/paper.Rmd
INITIAL EDITS PER P CONN SUGGESTS

Ice-associated seals rely on sea ice for a variety of activities, including breeding,
molting, pupping, and resting. In the Arctic, many of these activities occur in spring and
early summer (April-June) as sea ice begins to melt and retreat northward. Rapid
acceleration of climate change in Arctic ecosystems is thus of concern as the quantity and
quality of suitable habitat is forecast to decrease. In this paper, we use data collected
from satellite-linked telemetry tags deployed between 2005 and 2020 to investigate the
seasonal timing and environmental factors affecting sea ice use by seals (specifically,
bearded, ribbon, and spotted seals) in the Bering, and Chukchi seas. In addition to
providing baseline data on phenology, these data also allow us to refine availability
estimates needed to accurately estimate abundance from aerial survey counts of seals
basking on ice (i.e., to correct for the proportion of animals that are in the water while
surveys are conducted). Using generalized linear mixed pseudo-models to properly account
for temporal autocorrelation, we fit models with a variety of covariates (e.g.,
day-of-year, solar hour, age-sex class, wind speed, barometric pressure, temperature,
precipitation) to examine their ability to explain variation in hourly haul-out records.
We found evidence for strong diurnal and within season patterns in haul-out behavior, as
well as strong weather effects (particularly wind and temperature). In general, seals were
more likely to rest on ice in the middle of the day and when wind speed was low and
temperatures were higher. Haul-out propensity increased through March and April, peaking
in May and early June before declining again. The timing and frequency of haul-out events
also varied based on species and age-sex class. For ribbon and spotted seals, models with
year effects were highly supported, indicating that the timing and magnitude of haul-out
behavior varied among years. There was no support linking the variability in haul-out
timing to annual sea ice extent. Our analysis also emphasizes the importance of accounting
for weather factors affecting haul-out behavior when interpreting the number of seals
counted in aerial surveys.

@@

In this study, we use data from fifteen years of bio-logger records in a comprehensive
investigation of haul-out behavior by three ice-associated seals in the Bering and Chukchi
Seas. We focused our effort on bearded, spotted, and ribbon seals and do not include
ringed seals the analysis. Ringed seals are unique in their use of snow lairs in the
spring months and they warrant a separate study that includes additional environmental
covariates such as the onset of snow melt and snow depth. Our goals are threefold. First,
we wish to establish baseline estimates for the chronology of haul-out behavior in the
critical spring season for each species across different age and sex classes. Second, we
wish to refine estimates of haul-out availability corrections for aerial surveys in order
to improve estimates of seal abundance. Previously estimated availability correction
factors (e.g., [@bengtson2005pb], [@conn2014mee], and [@verhoef2014sm]) accounted for
variables such as the time-of-day and day-of-year, but did not investigate meteorological
variables that have been shown to influence haul-out behavior of walruses [@reder2003pb;
@udevitz2009cjz]. Further, previous correction factor analyses assumed that tagged seals
were representative of the population as a whole, even though the age and sex distribution
of telemetered individuals might differ from the population. Third, we explore the annual
variability in haul-out timing and possible linkage to dramatic changes in the
availability and extent of seasonal sea ice between 2005 and 2020. Our work extends the
scope of previous haul-out analyses, includes the influence of meteorological variability
and possible ramifications of non-representative age-sex samples, and investigates the
potential impact of a changing icescape on the behavior of these species.

@@

Haul-out behavior data from `r dplyr::n_distinct(deploy_tbl$speno)` bio-loggers deployed
on bearded, spotted, and ribbon seals was subset to include only records from 1 March to
15 July between `r min(lubridate::year(dat.sf$haulout_dt))` and
`r max(lubridate::year(dat.sf$haulout_dt))`. Bio-loggers were of the 'SPLASH' or 'SPOT'
family of tags developed by Wildlife Computers (Redmond, Washington, USA) and either
adhered to the hair on the seal or attached to the rear flipper inter-digital webbing. The
use of bio-loggers adhered to the back or head provide some benefits over flipper mounted
devices (e.g. increased satellite transmittal rates, locations at sea) but these are lost
during the annual molt and limits the effective length of haul-out timelines.
Additionally, sensors attached to the head or dorsal region are often dry while the seal
is at the surface and can slightly bias the percent-dry values recorded by the device. For
this study, in cases where both tag types were deployed, percent dry observations from the
flipper tag were usually preferred. Sex and age class were determined at the time of
deployment by experts in the field. Seals determined to be less than one year were
classified as young of the year. subadults are those seals likely greater than one year of
age but not yet sexually mature. Adults are likely sexually mature and older than
approximately four years. For those tags deployed on young of the year and transmitting
into the next year, the age class was advanced to subadult on March 1 of the following
year. Table \@ref(tab:tblDeploy) provides a summary of these deployments and data received
from them.

@@

```{r seal-hours-calendar, fig.height = 4.75, include=TRUE, fig.id = "dataCal", fig.cap="Distribution of of haul-out behavior data from 1 March to 15 July for each species. The higher density of data from May and June coincides with peak molting when seals (and their attached bio-loggers) are more likely hauled out. Additionally, a majority of deployments started in April and May. Data were collated across all years between 2005 and 2020."}
ggplot(data = seal_hours_tbl) + 
  geom_point(aes(x = wkofyr, y = dayofwk, color = n, size = n), stroke = 0, shape = 15) +
  scale_size(range = c(1,4.35)) +
  scale_color_gradient(low = "#6baed6", high = "#08306b") +
  scale_x_continuous(breaks = c(13,18,23,27),
                     labels = c("April", "May", "June", "July")) +
  guides(color = "none",
        size = guide_legend(override.aes = list(color = "#2171b5"),
                            title.position = "bottom", title.hjust = 0.5)
        ) +
  facet_grid(species ~ .) + 
  theme(axis.text.y = element_blank(),
        panel.grid = element_blank(),
        legend.position = "bottom") +
  labs(x = "", y = "day of week", color = "seal-hours",
       size = "seal hours of bio-logger data") +
  ggtitle("Haul-out behavior data is concentrated in the months of May and June",
          subtitle = "data presented in seal-hours collated across years (2005-2020)")
```

@@

Of key interest in this study is the relationship between haul-out behavior and a variety
of environmental covariates that vary with space and time. We explored use of a movement
model to predict locations at specific times. However, the sparse nature of data from some
tags, especially those mounted to the rear flipper, resulted in poor modeling performance
or convergence issues. For this study, we calculated a weighted average daily location
where the inverse of the estimated Argos location error was used for the weight. Any days
where haul-out observations were present without any location data were filled from the
last calculated daily location and any days where the location intersected with land were
removed from the dataset. Figure \@ref(fig:dataMap) shows the spatial distribution of
available haul-out behavior data for each species across the study area.

@@

Notably missing from this list of explanatory variables is any spatial-temporal
representation of sea ice concentration, area, or extent. This may seem counter-intuitive
when modeling the haul-out behavior of seal species with such a close association to sea
ice. Seals haul out in the presence of sea ice and we could assess the local concentration
of sea ice during these events. However, seals also spend time in the water away from sea
ice. Thus, the challenge is how best to associate that seal's at-sea behavior with a sea
ice concentration. Would it be the concentration at the last known haul-out event? The
next known haul-out event? Or, the closest sea ice concentration available? Without a
meaningful means to associate a sea ice covariate with the seal throughout the duration of
the observation record, such parameters are not viable for inclusion in this model.

We did explore whether the annual variation in spring sea ice extent in the Bering Sea
influences the peak timing of seal haul-out behavior. For this, all sea ice concentration
grid cells (25 km^2^) in the study area with greater than 15% concentration were summed to
get total sea ice extent for each day between 15 February and 15 July across all years
(`r min(lubridate::year(dat.sf$haulout_dt))`-`r max(lubridate::year(dat.sf$haulout_dt))`).
The maximum sea ice extent for each year was determined from this calculation. We obtained
sea ice concentration data from the Nimbus-7 SMMR and DMSP SSM/I-SSMIS Passive Microwave
Dataset, Version 1 [@cavalieri1996].

```{r tblCovariates, include=TRUE, tab.id = "tblCovariates", tab.cap = "Explanatory covariates used in analyses of binary haul-out records for ice-associated seals."}
covariate_tbl <- tibble::tribble(
  ~Covariate, ~Type, ~Source, ~Description,
  "Age-sex class",   "Categorical", "Field Assessment", 
  "Classes included young-of-year, subadult, adult male and adult female. Young-of-year tags surviving past ~Feb 1 of the next year were treated as subadult in subsequent years.",
  "Hour", "Continuous; Fourier basis", "Bio-logger", 
  "We modeled solar hour using 6 variables of a Fourier-series basis to allow circularity, flexibility, and dimension reduction ",
  "Day",   "Continuous", "Bio-logger", 
  "Linear, quadratic, and cubic effects of day-of-year were modeled",
  "Precip", "Continuous", "NARR", 
  "Convective precipitation (kg/m2)",
  "Pressure", "Continuos", "NARR", 
  "Atmospheric pressure at sea level (kPa)",
  "Temp", "Continuous", "NARR",
  "Air temperatures predicted at 2m above the earth’s surface",
  "Wind", "Continuous", "NARR", 
  "Northerly and easterly vector components for wind were downloaded and converted into a single wind speed via the Euclidean norm.",
  "Northing", "Continuous", "Bio-logger","For bearded seals a measurement of northing was caclulated as latitude divided by the mean latitude across all locations"
)

@@

We followed Ver Hoef et al. [-@verhoef2014sm] in using linear, quadratic, and cubic
effects of day-of-year to represent seasonal changes in behavior. However, unlike previous
models for harbor seals [@london2012] and ice-associated seals [@verhoef2014sm], which
treated hour-of-day as a 24-level categorical variable to capture diurnal cycles, we
adopted a continuous formulation based on Fourier series that provides a flexible model
while preserving the inherit circularity needed for time-of-day effects (i.e., hour 0
should be equal to hour 24). It also represents hour-of-day with 6 variables, which is a
considerable reduction when compared to a 24 variable model, especially when fitting
models with interactions between hour-of-day and other variables (e.g., age-sex class,
day-of-year). According to this approach, we used the following specification for
hour-of-day effects:

@@

For the second set of models examining inter-annual variation in sea ice use, we fitted
models with year by day-of-year interactions. However, in this case we only included
year$\times$day and year$\times$day2, omitting the main effects of year as well as
year$\times$day3 interactions as models with the latter effects were numerically unstable.
However, the modeled interactions were sufficient to allow shifts in haul-out
distribution, as one can show mathematically that a simple horizontal shift in timing of
haul-out distributions does not affect the main effects or cubic terms in a polynomial
regression model.

@@

A typical model fitting exercise would also include a model selection process. However,
AIC (and similar criteria) is not suitable when using pseudo-likelihoods. After fitting
GLMPM models, we used "type III" F-tests to calculate p-values [@verhoef2010cs] and rely
on those insights to evaluate model performance and important terms. We also produced
predictions of haul-out behavior as a function of influential predictors (e.g. solar hour,
day-of-year, age-sex). Weather covariates for these predictions were based on daily or
hourly (GAM) smoothed weather covariate values across the study region. Such predictions
were then used to develop haul-out probability surfaces, explore marginal effects of
weather covariates, and determine annual peaks in haul-out activity. The timing of peak
haul-out behavior was further used to regress against the annual maximum sea ice extent in
the study region.

>>>>

Ice-associated seals rely on sea ice for a variety of activities, including breeding,
molting, pupping, and resting. In the Arctic, many of these activities occur in spring and
early summer (April-June) as sea ice begins to melt and retreat northward. Rapid
acceleration of climate change in Arctic ecosystems is thus of concern as the quantity and
quality of suitable habitat is forecast to decrease. In this paper, we use data collected
from satellite-linked telemetry tags deployed between 2005 and 2020 to investigate the
seasonal timing and environmental factors affecting sea ice use by seals (specifically,
bearded, ribbon, and spotted seals) in the Bering, and Chukchi seas. In addition to
providing baseline data on phenology, these data also allow us to refine availability
estimates needed to accurately estimate abundance from aerial survey counts of seals
basking on ice (i.e., to correct for the proportion of animals that are in the water while
surveys are conducted). Using generalized linear mixed pseudo-models to properly account
for temporal autocorrelation, we fit models with a variety of covariates (e.g.,
day-of-year, solar hour, age-sex class, wind speed, barometric pressure, temperature,
precipitation) to examine their ability to explain variation in hourly haul-out records.
We found evidence for strong diurnal and within season patterns in haul-out behavior, as
well as strong weather effects (particularly wind and temperature). In general, seals were
more likely to rest on ice in the middle of the day and when wind speed was low and
temperatures were higher. Haul-out propensity increased through March and April, peaking
in May and early June before declining again. The timing and frequency of haul-out events
also varied based on species and age-sex class. For ribbon and spotted seals, models with
year effects were highly supported, indicating that the timing and magnitude of haul-out
behavior varied among years. However, we did not find evidence that haul-out
timing was linked annual sea ice extent. Our analysis emphasizes the importance of accounting
for variation in haul-out behavior when interpreting the number of seals
counted in aerial surveys.

@@

In this study, we use data from fifteen years of bio-logger records in a comprehensive
investigation of haul-out behavior by three ice-associated seals in the Bering and Chukchi
Seas. We focused our effort on bearded, spotted, and ribbon seals and do not include
ringed seals in the analysis. Ringed seals are unique in their use of snow lairs in the
spring months and they warrant a separate study that includes additional environmental
covariates such as the onset of snow melt and snow depth. Our goals are threefold. First,
we wish to establish baseline estimates for the chronology of haul-out behavior in the
critical spring season for each species across different age and sex classes. Second, we
wish to refine estimates of haul-out availability corrections for aerial surveys in order
to improve estimates of seal abundance. Previously estimated availability correction
factors (e.g., [@bengtson2005pb], [@conn2014mee], and [@verhoef2014sm]) accounted for
variables such as the time-of-day and day-of-year, but did not investigate meteorological
variables that have been shown to influence haul-out behavior of walruses [@reder2003pb;
@udevitz2009cjz]. Further, previous correction factor analyses assumed that tagged seals
were representative of the population as a whole, even though the age and sex distribution
of telemetered individuals might differ from the population. Third, we explore the annual
variability in haul-out timing and possible linkage to dramatic changes in the
availability and extent of seasonal sea ice between 2005 and 2020. Our work extends the
scope of previous haul-out analyses, includes the influence of meteorological variability
and possible ramifications of non-representative age-sex samples, and investigates the
potential impact of a changing icescape on the behavior of these species.

@@

Haul-out behavior data from `r dplyr::n_distinct(deploy_tbl$speno)` bio-loggers deployed
on bearded, spotted, and ribbon seals was subset to include only records from 1 March to
15 July between `r min(lubridate::year(dat.sf$haulout_dt))` and
`r max(lubridate::year(dat.sf$haulout_dt))`. Bio-loggers were of the 'SPLASH' or 'SPOT'
family of tags developed by Wildlife Computers (Redmond, Washington, USA) and either
adhered to the hair on the seal or attached to the rear flipper inter-digital webbing. The
use of bio-loggers adhered to the back or head provide some benefits over flipper mounted
devices (e.g. increased satellite transmittal rates, locations at sea) but these are lost
during the annual molt and limits the effective length of haul-out timelines.
Additionally, sensors attached to the head or dorsal region are often dry while the seal
is at the surface and can slightly bias the percent-dry values recorded by the device. For
this study, in cases where both tag types were deployed, percent dry observations from the
flipper tag were preferred. Sex and age class were determined at the time of
deployment by experts in the field. Seals determined to be less than one year were
classified as young of the year. subadults are those seals likely greater than one year of
age but not yet sexually mature. Adults are likely sexually mature and older than
approximately four years. For those tags deployed on young of the year and transmitting
into the next year, the age class was advanced to subadult on March 1 of the following
year. Table \@ref(tab:tblDeploy) provides a summary of these deployments and data received
from them.

@@

```{r seal-hours-calendar, fig.height = 4.75, include=TRUE, fig.id = "dataCal", fig.cap="Distribution of haul-out behavior data from 1 March to 15 July for each species. The higher density of data from May and June coincides with peak molting when seals (and their attached bio-loggers) are more likely hauled out. Additionally, a majority of deployments started in April and May. Data were collated across all years between 2005 and 2020."}
ggplot(data = seal_hours_tbl) + 
  geom_point(aes(x = wkofyr, y = dayofwk, color = n, size = n), stroke = 0, shape = 15) +
  scale_size(range = c(1,4.35)) +
  scale_color_gradient(low = "#6baed6", high = "#08306b") +
  scale_x_continuous(breaks = c(13,18,23,27),
                     labels = c("April", "May", "June", "July")) +
  guides(color = "none",
        size = guide_legend(override.aes = list(color = "#2171b5"),
                            title.position = "bottom", title.hjust = 0.5)
        ) +
  facet_grid(species ~ .) + 
  theme(axis.text.y = element_blank(),
        panel.grid = element_blank(),
        legend.position = "bottom") +
  labs(x = "", y = "day of week", color = "seal-hours",
       size = "seal hours of bio-logger data") +
  ggtitle("Haul-out behavior data is concentrated in the months of May and June",
          subtitle = "data presented in seal-hours collated across years (2005-2020)")
```

@@

Of key interest in this study is the relationship between haul-out behavior and a variety
of environmental covariates that vary with space and time. We explored the use of a movement
model to predict locations at specific times. However, the sparse nature of data from some
tags, especially those mounted to the rear flipper, resulted in poor modeling performance
or convergence issues. For this study, we calculated a weighted average daily location
where the inverse of the estimated Argos location error was used for the weight. Any days
where haul-out observations were present without any location data were filled from the
last calculated daily location and any days where the location intersected with land were
removed from the dataset. Figure \@ref(fig:dataMap) shows the spatial distribution of
available haul-out behavior data for each species across the study area.

@@

Notably missing from this list of explanatory variables is any spatial-temporal
representation of sea ice concentration, area, or extent. This may seem counterintuitive
when modeling the haul-out behavior of seal species with such a close association to sea
ice. Seals haul out in the presence of sea ice and we could assess the local concentration
of sea ice during these events. However, seals also spend time in the water away from sea
ice. Thus, the challenge is how best to associate that seal's at-sea behavior with a sea
ice concentration. Would it be the concentration at the last known haul-out event? The
next known haul-out event? Or, the closest sea ice concentration available? Without a
meaningful means to associate a sea ice covariate with the seal throughout the duration of
the observation record, such parameters are not viable for inclusion in this model.

We did explore whether the annual variation in maximum spring sea ice extent in the Bering
Sea influences the seasonal peak of seal haul-out behavior. In particular, we used sea ice
concentration data from the Nimbus-7 SMMR and DMSP SSM/I-SSMIS Passive Microwave Dataset,
Version 1 [@cavalieri1996] to calculate maximum sea ice extent. All sea ice concentration
grid cells (25 km^2^) in the study area with greater than 15% concentration were summed to
get total sea ice extent for each day between 15 February and 15 July across all years.
Maximum spring sea ice extent was simply the largest daily value for each year.

```{r tblCovariates, include=TRUE, tab.id = "tblCovariates", tab.cap = "Explanatory covariates used in analyses of binary haul-out records for ice-associated seals."}
covariate_tbl <- tibble::tribble(
  ~Covariate, ~Type, ~Source, ~Description,
  "Age-sex class",   "Categorical", "Field Assessment", 
  "Classes included young-of-year, subadult, adult male and adult female. Young-of-year tags surviving past ~Feb 1 of the next year were treated as subadult in subsequent years.",
  "Hour", "Continuous; Fourier basis", "Bio-logger", 
  "We modeled solar hour using 6 variables of a Fourier-series basis to allow circularity, flexibility, and dimension reduction ",
  "Day",   "Continuous", "Bio-logger", 
  "Linear, quadratic, and cubic effects of day-of-year were modeled",
  "Precip", "Continuous", "NARR", 
  "Convective precipitation (kg/m2)",
  "Pressure", "Continuous", "NARR", 
  "Atmospheric pressure at sea level (kPa)",
  "Temp", "Continuous", "NARR",
  "Air temperatures predicted at 2m above the earth’s surface",
  "Wind", "Continuous", "NARR", 
  "Northerly and easterly vector components for wind were downloaded and converted into a single wind speed via the Euclidean norm.",
  "Northing", "Continuous", "Bio-logger","For bearded seals a measurement of northing was caclulated as latitude divided by the mean latitude across all locations"
)

@@

We followed Ver Hoef et al. [-@verhoef2014sm] in using linear, quadratic, and cubic
effects of day-of-year to represent seasonal changes in behavior. However, unlike previous
models for harbor seals [@london2012] and ice-associated seals [@verhoef2014sm], which
treated hour-of-day as a 24-level categorical variable to capture diurnal cycles, we
adopted a continuous formulation based on Fourier series that provides a flexible model
while preserving the inherent circularity needed for time-of-day effects (i.e., hour 0
should be equal to hour 24). It also represents hour-of-day with 6 variables, which is a
considerable reduction when compared to a 24 variable model, especially when fitting
models with interactions between hour-of-day and other variables (e.g., age-sex class,
day-of-year). According to this approach, we used the following specification for
hour-of-day effects:

@@

For the second set of models examining inter-annual variation in sea ice use, we fitted
models with year by day-of-year interactions. However, in this case we only included
year$\times$day and year$\times$day2, omitting the main effects of year as well as
year$\times$day3 interactions because models with the latter effects were numerically unstable.
However, the modeled interactions were sufficient to allow shifts in haul-out
distribution, as one can show mathematically that a simple horizontal shift in timing of
haul-out distributions does not affect the main effects or cubic terms in a polynomial
regression model.

@@

A typical model fitting exercise would also include a model selection process. However,
AIC (and similar criteria) is not suitable when using pseudo-likelihoods. After fitting
GLMPM models, we used "type III" F-tests to calculate p-values [@verhoef2010cs] to
evaluate model performance and important terms. We also produced predictions of haul-out
behavior as a function of influential predictors (e.g. solar hour, day-of-year, age-sex).
Weather covariates for these predictions were based on daily or hourly smoothed weather
covariate values across the study region. Such predictions were then used to develop
haul-out probability surfaces, explore marginal effects of weather covariates, and
determine annual peaks in haul-out activity. The timing of peak haul-out behavior was
further used to regress against the annual maximum sea ice extent in the study region.
